A Partitioning Algorithm for Maximum Common Subgraph Problems\thanks{This work was supported by a pile of money from a funding agency.}

### 2017-02-16 paper/paper.tex
Avoid big spaces between words
cref

@@

We initially assume that graphs are unlabelled, undirected and without loops
(Section~\ref{sec:extensions} describes how these restrictions may be relaxed).
The vertices and edge sets of a graph $G$ are denoted $\V(G)$ and $\E(G)$.  The
set of vertices adjacent to vertex $v$ in graph $G$ is called the
\emph{neighbourhood} of $v$, denoted $\N(G, v)$. We denote by $\invN(G, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $G$ is a graph
consisting of some of the vertices of $G$, and all of the edges between these
vertices (in this paper all subgraphs are induced). A \emph{common subgraph} of
two graphs is a graph which is (isomorphic to) a subgraph of two graphs
simultaneously, and a \emph{maximum} common subgraph is one with as many
vertices as possible.

@@

With these definitions established, we now present \McSplit, a branch and bound
algorithm which uses a new vertex-labelling scheme to avoid explicitly
maintaining the set of feasible mappings for each vertex in $\V(G)$.
\iffalse
As we will explore in
Section~\ref{sec:comparison}, the algorithm may be viewed as a version of the
constraint programming forward-checking (CP-FC) algorithm using a compact data
structure that exploits the nature of the MCS problem.  It is sufficiently
different, however, that it is helpful to describe the algorithm without
reference to previous approaches.

@@

We will use the graphs $G$ and $H$ in Figure~\ref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; one
example is the mapping $\{1a, 2f, 3d, 5b\}$ where vertex $1$ is assigned to
vertex $a$, $2$ is assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

We begin by mapping vertex $1$ to vertex $a$, giving $M=\{1a\}$.  Now label
each unmatched vertex according to whether it is adjacent to vertex $1$ (in
$G$) or vertex $a$ (in $H$), as shown in Figure~\ref{fig:alg2}.  Adjacent
vertices have label 1; non-adjacent vertices have label 0.  We can extend $M$
with a vertex mapping $vw$, with $v \in \V(G)$ and $w \in \V(H)$, if and only
if $v$ and $w$ have the same label.  This property, that two vertices may be
mapped together if and only if they share a label, is the algorithm's main
invariant.

@@

    $\{1a\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $2$ & 0 \\
        $3$ & 0 \\
        $4$ & 1 \\
        $5$ & 1 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 1 \\
        $c$ & 1 \\
        $d$ & 0 \\
        $e$ & 1 \\
        $f$ & 0 \\
    \hline
    \end{tabular}
\end{minipage}
    \caption{Labels on vertices after mapping $1$ to $a$}
\label{fig:alg2}
\end{figure}

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (Figure~\ref{fig:alg3}).  Each unmapped vertex $v \in \V(G)$ is labelled
with a two-character bit string, indicating its adjacency to each of
the two mapped vertices in $\V(G)$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to vertex $1$ but is adjacent
to vertex $2$.  Labels are given to unmapped vertices in $\V(H)$ in a similar fashion,
showing adjacency to matched vertices $a$ and $d$.

Our invariant is maintained: we can extend $M$ by a vertex pairing if and only
if the two vertices have the same label. Note also that the labels can be
calculated efficiently, since each label (in Figure~\ref{fig:alg3}) is prefixed
by the label at the previous level of the search tree (Figure~\ref{fig:alg2}).

@@

    $\{1a, 2d\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $3$ & 01 \\
        $4$ & 10 \\
        $5$ & 11 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 11 \\
        $c$ & 11 \\
        $e$ & 10 \\
        $f$ & 01 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $2$ to $d$}
\label{fig:alg3}
\end{figure}

The algorithm backtracks when the incumbent, i.e. the largest matching found so far, is at least as large
as a calculated bound given $M$ and the current labelling. To demonstrate how
this bound is calculated, we consider the situation one level deeper in the
search tree shown in Figure~\ref{fig:alg4}.

@@

    $\{1a, 2d, 3f\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $4$ & 100 \\
        $5$ & 111 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 111 \\
        $c$ & 111 \\
        $e$ & 101 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $3$ to $f$}
\label{fig:alg4}
\end{figure}

@@

\paragraph{Algorithm \ref{McCliqueAlg} Explained} The recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

@@

The correctness of \McSplit\ can be seen by noting its equivalence to CP-FC
\citep{DBLP:conf/cp/NdiayeS11}, if we replace the label-class representation of
$\mathit{future}$ with domain stores in Algorithm \ref{McCliqueAlg}.  To obtain
the CP-FC algorithm, replace \linerangeref{InnerLoop}{InnerLoopEnd} of
Algorithm \ref{McCliqueAlg} with Algorithm \ref{cpAlg}.  The pruning of domains
carried out in these lines can be easily seen to be equivalent to the
corresponding lines in Algorithm \ref{McCliqueAlg}.

@@

\paragraph{Directed graphs without edge labels} Let $A_G$ and $A_H$ be
adjacency matrices of $G$ and $H$. For each vertex pair $(t,u)$ in graph $G$ or
$H$, the adjacency matrix entry takes the value 0 if the vertices are not
adjacent, 1 if the two vertices share a single edge in the direction $t
\rightarrow u$, 2 if they share a single edge in the direction $u \rightarrow
t$, and 3 if there are edges in both directions. Where
\linerangeref{NewPWithEdge}{InnerLoopEnd} of the basic algorithm split the
label class $\langle P',T' \rangle$ in two, we now perform a four-way split
where each vertex is classified according to the label on its adjacency matrix
entry from $v$ or $w$.  This is shown in Algorithm~\ref{labDirAlg}, where
$L=\{0,1,2,3\}$.

@@

\paragraph{Undirected with edge labels} Each adjacency matrix
entry contains an edge label, or a null entry (non-adjacent) $0$.
We use Algorithm~\ref{labDirAlg}, by letting $L$ be
the union of $\{0\}$ with the set of all labels that appear in the input
graphs. Since there may be up to $g + h$ distinct labels, the loop in
Algorithm~\ref{labDirAlg} may execute up to $g + h$ times and
requires $\BigO{n^2}$ time per search
node.  It is possible to modify the algorithms to run in $\BigO{n \log n}$ per
search node.  First, run lines 17-19
of Algorithm~\ref{McCliqueAlg} to create a new label-class of vertices that are not
adjacent to $v$ or $w$, and remove these vertices from $\langle P',T' \rangle$.
Next, sort $P'$ and $T'$ in ascending order of the label on the edge from $v$
or $w$ to each vertex. We can then create the label classes corresponding to
each edge label by simultaneously traversing $P'$ and $T'$ from left to right.

@@

\paragraph{Top-Down Strategy} I'VE RE-ADDED A SHORTER VERSION OF THIS
PARAGRAPH, WHICH WAS COMMENTED OUT BY PATRICK. IT NEEDS WORK, BUT WE NEED TO
EXPLAIN THE SIP EXPERIMENT. We can modify the \McSplit\ algorithm to use a
top-down strategy based on \citet{UpcomingAAAIPaper} by calling the main
\FuncSty{McSplit} method once per goal size ($g, g-1, g-2, \dots$).  We
backtrack (\lineref{PruneSearch} of Algorithm~\ref{McCliqueAlg}) when the bound is
strictly less than the goal size, and terminate when a solution of the goal
size is found.

>>>>

% cref style
\crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

@@

We initially assume that graphs are unlabelled, undirected and without loops
(\cref{sec:extensions} describes how these restrictions may be relaxed).
The vertices and edge sets of a graph $G$ are denoted $\V(G)$ and $\E(G)$.  The
set of vertices adjacent to vertex $v$ in graph $G$ is called the
\emph{neighbourhood} of $v$, denoted $\N(G, v)$. We denote by $\invN(G, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $G$ is a graph
consisting of some of the vertices of $G$, and all of the edges between these
vertices (in this paper all subgraphs are induced). A \emph{common subgraph} of
two graphs is a graph which is (isomorphic to) a subgraph of two graphs
simultaneously, and a \emph{maximum} common subgraph is one with as many
vertices as possible.

@@

With these definitions established, we now present \McSplit, a branch and bound
algorithm which uses a new vertex-labelling scheme to avoid explicitly
maintaining the set of feasible mappings for each vertex in $\V(G)$.
\iffalse
As we will explore in
\cref{sec:comparison}, the algorithm may be viewed as a version of the
constraint programming forward-checking (CP-FC) algorithm using a compact data
structure that exploits the nature of the MCS problem.  It is sufficiently
different, however, that it is helpful to describe the algorithm without
reference to previous approaches.

@@

We will use the graphs $G$ and $H$ in \cref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; one
example is the mapping $\{1a, 2f, 3d, 5b\}$ where vertex $1$ is assigned to
vertex $a$, $2$ is assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

We begin by mapping vertex $1$ to vertex $a$, giving $M=\{1a\}$.  Now label
each unmatched vertex according to whether it is adjacent to vertex $1$ (in
$G$) or vertex $a$ (in $H$), as shown in \cref{fig:alg2}.  Adjacent
vertices have label 1; non-adjacent vertices have label 0.  We can extend $M$
with a vertex mapping $vw$, with $v \in \V(G)$ and $w \in \V(H)$, if and only
if $v$ and $w$ have the same label.  This property, that two vertices may be
mapped together if and only if they share a label, is the algorithm's main
invariant.

@@

    $\{1a\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $2$ & 0 \\
        $3$ & 0 \\
        $4$ & 1 \\
        $5$ & 1 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 1 \\
        $c$ & 1 \\
        $d$ & 0 \\
        $e$ & 1 \\
        $f$ & 0 \\
    \hline
    \end{tabular}
\end{minipage}
    \caption{Labels on vertices after mapping $1$ to $a$}
\label{fig:alg2}
\end{figure}

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (\cref{fig:alg3}).  Each unmapped vertex $v \in \V(G)$ is labelled
with a two-character bit string, indicating its adjacency to each of
the two mapped vertices in $\V(G)$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to vertex $1$ but is adjacent
to vertex $2$.  Labels are given to unmapped vertices in $\V(H)$ in a similar fashion,
showing adjacency to matched vertices $a$ and $d$.

Our invariant is maintained: we can extend $M$ by a vertex pairing if and only
if the two vertices have the same label. Note also that the labels can be
calculated efficiently, since each label (in \cref{fig:alg3}) is prefixed
by the label at the previous level of the search tree (\cref{fig:alg2}).

@@

    $\{1a, 2d\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $3$ & 01 \\
        $4$ & 10 \\
        $5$ & 11 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 11 \\
        $c$ & 11 \\
        $e$ & 10 \\
        $f$ & 01 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $2$ to $d$}
\label{fig:alg3}
\end{figure}

The algorithm backtracks when the incumbent, i.e. the largest matching found so far, is at least as large
as a calculated bound given $M$ and the current labelling. To demonstrate how
this bound is calculated, we consider the situation one level deeper in the
search tree shown in \cref{fig:alg4}.

@@

    $\{1a, 2d, 3f\}$
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $G$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $4$ & 100 \\
        $5$ & 111 \\
    \hline
    \end{tabular}
\end{minipage}
\quad
\begin{minipage}[t]{0.3\linewidth}
    \centering Labelling of $H$
    \begin{tabular}[t]{cc}
    \hline
        Vertex & Label\\
    \hline
        $b$ & 111 \\
        $c$ & 111 \\
        $e$ & 101 \\
    \hline
    \end{tabular}
\end{minipage}
\caption{After mapping $3$ to $f$}
\label{fig:alg4}
\end{figure}

@@

\paragraph{\cref{McCliqueAlg} Explained} The recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

@@

The correctness of \McSplit\ can be seen by noting its equivalence to CP-FC
\citep{DBLP:conf/cp/NdiayeS11}, if we replace the label-class representation of
$\mathit{future}$ with domain stores in \cref{McCliqueAlg}.  To obtain
the CP-FC algorithm, replace \linerangeref{InnerLoop}{InnerLoopEnd} of
\cref{McCliqueAlg} with \cref{cpAlg}.  The pruning of domains
carried out in these lines can be easily seen to be equivalent to the
corresponding lines in \cref{McCliqueAlg}.

@@

\paragraph{Directed graphs without edge labels} Let $A_G$ and $A_H$ be
adjacency matrices of $G$ and $H$. For each vertex pair $(t,u)$ in graph $G$ or
$H$, the adjacency matrix entry takes the value 0 if the vertices are not
adjacent, 1 if the two vertices share a single edge in the direction $t
\rightarrow u$, 2 if they share a single edge in the direction $u \rightarrow
t$, and 3 if there are edges in both directions. Where
\linerangeref{NewPWithEdge}{InnerLoopEnd} of the basic algorithm split the
label class $\langle P',T' \rangle$ in two, we now perform a four-way split
where each vertex is classified according to the label on its adjacency matrix
entry from $v$ or $w$.  This is shown in \cref{labDirAlg}, where
$L=\{0,1,2,3\}$.

@@

\paragraph{Undirected with edge labels} Each adjacency matrix
entry contains an edge label, or a null entry (non-adjacent) $0$.
We use \cref{labDirAlg}, by letting $L$ be
the union of $\{0\}$ with the set of all labels that appear in the input
graphs. Since there may be up to $g + h$ distinct labels, the loop in
\cref{labDirAlg} may execute up to $g + h$ times and
requires $\BigO{n^2}$ time per search
node.  It is possible to modify the algorithms to run in $\BigO{n \log n}$ per
search node.  First, run lines 17-19
of \cref{McCliqueAlg} to create a new label-class of vertices that are not
adjacent to $v$ or $w$, and remove these vertices from $\langle P',T' \rangle$.
Next, sort $P'$ and $T'$ in ascending order of the label on the edge from $v$
or $w$ to each vertex. We can then create the label classes corresponding to
each edge label by simultaneously traversing $P'$ and $T'$ from left to right.

@@

\paragraph{Top-Down Strategy} I'VE RE-ADDED A SHORTER VERSION OF THIS
PARAGRAPH, WHICH WAS COMMENTED OUT BY PATRICK. IT NEEDS WORK, BUT WE NEED TO
EXPLAIN THE SIP EXPERIMENT. We can modify the \McSplit\ algorithm to use a
top-down strategy based on \citet{UpcomingAAAIPaper} by calling the main
\FuncSty{McSplit} method once per goal size ($g, g-1, g-2, \dots$).  We
backtrack (\lineref{PruneSearch} of \cref{McCliqueAlg}) when the bound is
strictly less than the goal size, and terminate when a solution of the goal
size is found.
